% chapter 05
% include mechanism/switching model figure here
\chapter{Conclusions and Future Directions}\label{ch:conclusions}

What is the molecular mechanism of Ena/VASP on single and bundled filaments?
How do bundling proteins effect other actin binding proteins?
What are the effect of WH2 containing molecules?

\section{Ena/VASP mechanisms}\label{ena-mechanism-conclusions}
Understanding the mechanism of formins and their regulation by protein binding partners as well as outside forces has taken a step forward in the last few years. Though the mechanistic understanding of Ena/VASP lags behind that of formins, great progress is being made in understanding Ena/VASP's processive mechanism and how it functions within cells. There has been initial insight in how these two processive polymerases work together within cells, but much more needs to be done to understand their collaboration and concurrent regulation. These processive polymerases are mechanistically interesting to compare and contrast to understand how different proteins maintain actin filament contact as well as how they are regulated within the cell. Understanding their individual functions as well as how they work together will give a clearer picture of what is happening at filopodia and the leading edge of the cell. 

\section{Ena/VASP and bundled actin filaments}\label{ena-bundles-conclusions}
Ena/VASP making filopodia:
mechanism with convergent elongation

Different actin networks, bundling protein can regulate
Fascin specific for filopodia

\section{Actin bundling proteins and their effect on other ABPs}\label{bundler-conclusions}

As we've seen with Ena, bundling proteins 
Effect on actin nucleation: fascin slower? Or not measured so shouldn't mention

\section{Pathogenic bacteria target the host actin cytoskeleton}\label{bacteria-conclusions}
Actin is a good target
Mechanism of ACD oligomers
Interesting that VopL/F also fall victim to ACD
Why VopL/F works, differences between with and without monomer free

\section{Concluding Remarks}\label{final-conclusions}
Looked at ABPs